~\Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores

### 2024-09-26 manuscript/paper.tex
changes caused by knitting the Rmd

Ecological release theory suggests that a population's niche should
change when important species interactions (e.g., resource competition)
are relaxed or removed (reviewed in Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmannetal2021}{2021})). Removal is posited to
create ecological opportunities---i.e., aspects of the niche become
newly accessible, and the focal population shifts and/or expands its
resource use to the new opportunity (Parent and Crespi
(\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})). Ecological
release may then be followed by adaptive morphological evolution (Parent
and Crespi (\protect\hyperlink{ref-ParentandCrespi2009}{2009});
Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmannetal2021}{2021})). For example, a
population undergoing ecological release can experience disruptive
selection on males and females stemming from intraspecific competition
over newly accessible resources (Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Bolnick and Lau
(\protect\hyperlink{ref-BolnickandLau2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011})). This is
predicted to result in intersexual divergence in habitat use and
associated phenotypes (Schoener
(\protect\hyperlink{ref-Schoener1968}{1968}); Shine
(\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Butler and Losos
(\protect\hyperlink{ref-Butleretal2007}{2007}); Bolnick and Lau
(\protect\hyperlink{ref-BolnickandLau2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011}); but see Stuart
et al. (\protect\hyperlink{ref-Stuartetal2021}{2021}); Blain
(\protect\hyperlink{ref-Blain2022}{2022})). Such competition-driven
``intraspecific character displacement'' between the sexes is therefore
one explanation for the evolution of sexual dimorphism (Pfennig and
Pfennig (\protect\hyperlink{ref-PfennigandPfennig2012}{2012}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2015}{2015}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2017}{2017}), De Lisle,
Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).

Much of the theory and empirical data for character displacement between
the sexes is based on release from resource competition specifically
(e.g., Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Cooper, Gilman,
and Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011}); Pfennig and
Pfennig (\protect\hyperlink{ref-PfennigandPfennig2012}{2012}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2015}{2015}); De Lisle,
Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).
However, release from predation should also result in ecological release
followed by the evolution of increased sexual dimorphism because the
absence of predators should generate ecological opportunity (Reimchen
and Nosil (\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Parent
and Crespi (\protect\hyperlink{ref-ParentandCrespi2009}{2009});
Herrmann, Stroud, and Losos
(\protect\hyperlink{ref-Herrmannetal2021}{2021})).

Here, we tested the prediction that a release from predation should
result in the evolution of increased sexual dimorphism using a
well-preserved, finely-resolved sequence of a fossil threespine
stickleback fish (Gasterosteus doryssus). The sequence in this
depositional environment is comprised of two lineages (Bell
(\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage I was a
low-armor form with zero to one dorsal spines and highly reduced
pelvises, on average. It lasted for at least 93,000 years before it was
replaced suddenly by a second lineage (Bell, Baumgartner, and Oslon
(\protect\hyperlink{ref-Belletal1985}{1985})). Lineage II appeared in
the depositional environment fully armored, with complete pelvic
girdles, two pelvic spines, and three dorsal spines, on average (Bell,
Travis, and Blouw (\protect\hyperlink{ref-Belletal2006}{2006}); Stuart,
Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020})). It
then immediately began evolving adaptive reduction in its armor traits
(Hunt, Bell, and Travis (\protect\hyperlink{ref-Huntetal2008}{2008});
Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020}))
until it reached the same low-armor state previously held by lineage I.
The observation of 93,000 years of low-armor stasis in lineage I (Bell,
Baumgartner, and Oslon (\protect\hyperlink{ref-Belletal1985}{1985})) and
the observation of rapid evolution of armor loss by lineage II, both
suggest that this depositional environment lacked piscivorous fish
(e.g., trout and other salmonids known to prey on modern threespine
stickleback). Armor maintenance in extant threespine stickleback
(Gasterosteus aculeatus) correlates strongly the presence of vertebrate
piscivores; populations with less predation pressure typically have less
armor (Reimchen (\protect\hyperlink{ref-Reimchen1994}{1994}), Reimchen
and Nosil (\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Bell et
al. (\protect\hyperlink{ref-Belletal1993}{1993}); Roesti et al.
(\protect\hyperlink{ref-Roestietal2023}{2023})). Consistently, only
three fossil trout have been recorded from this paleolake basin, sampled
from the same rock that has revealed \textgreater15,000 threespine
stickleback fossils as well as occasional killifish (Fundulus
nevadensis) (Bell (\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni,
Bell, and Stuart (\protect\hyperlink{ref-Cerasonietal2024}{2024})).
Finally, tooth wear data from the lineage II sequence suggest that
individuals in this lineage began eating more planktonic prey over time,
expanding toward an open-water niche from the benthic, bottom-feeding
niche they started with (Purnell et al.~2007). That this expansion into
open water by lineage II coincided with armor reduction is further
indicative of a limnetic system with fewer salmonid piscivores (Schluter
and McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992});
Vamosi and Schluter
(\protect\hyperlink{ref-VamosiandSchluter2004}{2004}); Roesti et al.
(\protect\hyperlink{ref-Roestietal2023}{2023})). In sum, we infer that
lineage II likely started in a nearby paleolake basin that had predators
since it was armored when it arrived (Bell
(\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage II then
experienced release from predators and dietary expansion in the focal
paleolake basin, generating a scenario consistent with the starting
conditions for models predicting the evolution of increased sexual
dimorphism following ecological release.

@@

For our study, we inferred sex in G. doryssus fossils by comparing
multivariate morphological trait data from lineage II samples to the
same multivariate trait set collected from multiple populations in the
closely-related, extant Threespine Stickleback species complex
(Gasterosteus aculeatus). Crucially, we determined G. aculeatus sex
directly via dissection and/or PCR genotyping. This enabled us to use
Multiple Imputation by Chained Equations (MICE; Buuren and
Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) to build a
predictive multiple imputation algorithm (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) based on the
multivariate morphology of G. aculeatus individuals of known sex. We
applied this algorithm to the fossil data to impute individual sex based
on morphology, treating fossil sex as a missing variable. With sex
assigned to fossil specimens (and accounting for uncertainty in those
assignments), we fit a modified Ornstein-Uhlenbeck (OU) (Uhlenbeck and
Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model using a
Bayesian framework to test for evolution of sexual dimorphism in each
trait over \textasciitilde16,000 years of lineage II. We hypothesized
that sexual dimorphism should increase following ecological release from
predators.

@@

We used Gasterosteus doryssus data that were previously reported by
Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020}),
Voje, Bell, and Stuart (\protect\hyperlink{ref-Vojeetal2022}{2022}), and
Siddiqui et al. (\protect\hyperlink{ref-Siddiquietal2024}{2024}).
Briefly, the data were collected from fossil Series K from Quarry D
(Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})), dug from an open pit
diatomite mine at 9.526° N, 119.094° W, near Fernley, Nevada, USA.
Series K consisted of 18 samples taken at \textasciitilde1000-year
intervals, and mean sample times span \textasciitilde16,363 years. Fish
from series K were measured for 16 ecomorphological traits related to
armor, swimming, and feeding (Table 1). Series K started at a previously
documented horizon when a low armored lineage of stickleback with zero
to one dorsal spines, zero pelvic spines, and highly reduced pelvises
(i.e., lineage I described above) was completely replaced by a high
armored lineage of stickleback with three dorsal spines, two pelvic
spines, and complete pelvises (i.e., lineage II described above) (Bell,
Travis, and Blouw (\protect\hyperlink{ref-Belletal2006}{2006}); Bell
(\protect\hyperlink{ref-Bell2009}{2009}); Stuart, Travis, and Bell
(\protect\hyperlink{ref-Stuartetal2020}{2020})). Lineage II subsequently
evolved reduction in armor, body size, and traits related to swimming
and feeding (Bell, Travis, and Blouw
(\protect\hyperlink{ref-Belletal2006}{2006}); Stuart, Travis, and Bell
(\protect\hyperlink{ref-Stuartetal2020}{2020}); Siddiqui et al.
(\protect\hyperlink{ref-Siddiquietal2024}{2024})). The tempo and mode of
armor reduction during this sequence suggests adaptive evolution by
natural selection (Hunt, Bell, and Travis
(\protect\hyperlink{ref-Huntetal2008}{2008})), and we focus on the
multivariate evolution of sexual dimorphism by this second lineage.

@@

To span the gamut of stickleback diversity for our predictive imputation
model, we sampled modern stickleback from lakes containing generalist
stickleback populations (Hendry et al.
(\protect\hyperlink{ref-Hendryetal2009}{2009}); Bolnick
(\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes containing
benthic-limnetic species pairs (Baumgartner, Bell, and Weinberg
(\protect\hyperlink{ref-Baumgartneretal1988}{1988}); Schluter and
McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992})). The
generalist populations used here were the remainder of collections made
by YES in 2013 and were previously described in Stuart et al.
(\protect\hyperlink{ref-Stuartetal2017}{2017}) (Table S1). These samples
were fixed in formalin, then stained for bone with Alizarin Red in 2013.
Benthic and limnetic specimens were kindly loaned by D. Schluter and his
lab at University of British Columbia. They collected benthic and
limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little
Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos
specimens had been fixed whole in formalin and stored in 40\%
isopropanol. The specimens from the other lakes were initially preserved
whole in 95\% ethanol in the field before being gradually transferred to
water then formalin in the lab and ultimately stored in 40\%
isopropanol. In 2019, we stained these specimens for bone using Alizarin
Red. We next replicated fossil data collection (Table 1) on these extant
specimens. Standard length as well as pelvic-spine length on each side
were measured with calipers. We used a dissection microscope to count
dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right
and left-side pelvic girdle lengths and ectocoracoid lengths were
measured from ventral photographs taken using a Canon EOS Rebel T7 with
a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy
stand. Specimens were held in place for ventral photographs using a
small tabletop vise with an attached scale bar. Lateral X-rays were used
to measure dorsal spine length, number of pterygiophores anterior to the
pterygiophore holding the third spine, length of the pterygiophore just
anterior to the third spine, cleithrum length, and pre-maxilla ascending
branch length. We also counted vertebrae from the X-rays: abdominal
vertebrae were counted anterior to the first vertebra with a haemal
spine contacting an anal fin pterygiophore. Caudal vertebrae were
posterior, including the first vertebra with the haemal spine contacting
the anal fin pterygiophore (following Aguirre, Walker, and Gideon
(\protect\hyperlink{ref-Aguirreetal2014}{2014})). X-rays were taken with
an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the
Field Museum of Natural History. Specimens were exposed at 35kV and 4mA.
Small fish were exposed for 7s, medium fish for 8s, and large fish for
10s. We developed the film and scanned individual images of each fish
using the B\&W Negatives setting on an Epson Perfection 4990 Photo
flatbed at 2400 dpi. Measurements from photographs and X-rays were taken
with FIJI (Schindelin et al.
(\protect\hyperlink{ref-Schindelinetal2012}{2012})) and its plugin
ObjectJ (\url{https://sils.fnwi.uva.nl/bcb/objectj/}).

@@

To check for outliers, we calculated within-group means and standard
deviations for each trait separately for K series fossil specimens
(pooled across samples) and for extant specimens (within generalist,
benthic, or limnetic categories). We noted trait values greater than 3.0
standard deviations from the mean as potential outliers. We deemed 3.0
s.d. to be a reasonable threshold for detecting errors without excluding
biologically relevant values. We checked whether these potential
outliers were a result of data entry and collection error and corrected
them if they were. We turned the remaining outlier trait values to NAs.
We size-corrected continuous traits only, as they varied with size,
unlike count traits that are fixed during early development. We
regressed each trait on standard length using a mixed-model regression,
pooling all specimens, following Stuart et al.
(\protect\hyperlink{ref-Stuartetal2017}{2017}). We appended the size
corrected data to the uncorrected trait data frame and used all of our
data to build the imputation model described next.

@@

We used multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils many times. Imputations were performed using the MICE
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
algorithm and implemented in R (Team
(\protect\hyperlink{ref-R2022language}{2022})) with M = 100 imputations
performed. After imputing the sex of the fossil data, each completed
data set is then used to fit a modified Ornstein--Uhlenbeck (OU)
(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model
using a Bayesian framework to look for sexual dimorphism across a
variety of stickleback phenotypes.

@@

We impute the missing sex for the fossil data by sampling from the
posterior predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
of imputed data sets if the data users intend on performing Bayesians
analysis after imputation, which in this case, we do. Therefore, the
imputation algorithm is run to obtain a total of \(M = 100\) completed
datasets. In addition to this, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) suggests rather than
using Rubin's combining rules to combine across imputed data sets,
instead pool all of the draws from the posterior distributions across
all of the imputed data sets to estimate the posterior distributions of
parameters of interest. We proceed with our Bayesian analysis in this
manner.

\textcolor{red}{Akhil's stuff about validating the imputation model goes here.}

\hypertarget{completed-data-model}{%
\subsection{Completed Data Model}\label{completed-data-model}}

\textcolor{red}{We should note somewhere that we are only using the $W_{fossil}$ in the modeling part.  We drop the $W_{extant}$.  So just note that $W_{ti}$ is really $W_{fossil,ti}$.  Not sure how to say this, but we need to make it clear that we are only using the fossil data for the OU modeling.  Greg, check and see if this makes sense.}

\textcolor{red}{I think we need to move a lot of these details to the appendix.  }

For a given imputed dataset, let \(W_{ij}\) be the imputed sex and
\(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
\(j = 1,\ldots,n_{t}\). Note, we are inputting only the fossil data into
our model; connecting to the previous section, \(W_{ij}\) and
\(\boldsymbol{Y}_{ij}\) can be interpretted as \(W_{fossil,ij}\) and
\(\boldsymbol{Y}_{fossil,ij}\). In addition, we denote \(Y_{K,ij}\), the
last variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of
the fish, \begin{align}
{Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

@@

Because we are fitting a dataset with a stochastic structure on the
means of the phenotypes, and we are interested in the distribution of
the overall structure of these means, we analyze the data via a Bayesian
analysis. Bayesian data analysis is also more naturally used when we
have to impute data (\textbf{CITATION}). To aid in the sampling
procedure for the discrete phenotypes, we declare

>>>>

Ecological release theory suggests that a population's niche should
change when important species interactions like resource competition or
predation are relaxed or removed (reviewed in Herrmann, Stroud, and
Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})). Removal is
posited to create ecological opportunity---i.e., aspects of the niche
become newly accessible, and the focal population shifts and/or expands
its resource use new resources (Parent and Crespi
(\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})). Ecological
release may then be followed by adaptive morphological evolution as
traits change to reflect the new niche (Parent and Crespi
(\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})).

For example, a population undergoing ecological release can experience
disruptive selection on males and females stemming from intraspecific
competition over newly accessible resources (Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Bolnick and Lau
(\protect\hyperlink{ref-BolnickandLau2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011})). This is
predicted to result in intersexual divergence in habitat use and
associated phenotypes (Schoener
(\protect\hyperlink{ref-Schoener1968}{1968}); Shine
(\protect\hyperlink{ref-Shine1989}{1989}); Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Butler and Losos
(\protect\hyperlink{ref-Butleretal2007}{2007}); Bolnick and Lau
(\protect\hyperlink{ref-BolnickandLau2008}{2008}); Cooper, Gilman, and
Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011}); but see Stuart
et al. (\protect\hyperlink{ref-Stuartetal2021}{2021}); Blain
(\protect\hyperlink{ref-Blain2022}{2022})). Such competition-driven
``intraspecific character displacement'' between the sexes is therefore
one explanation for the evolution of sexual dimorphism (Pfennig and
Pfennig (\protect\hyperlink{ref-PfennigandPfennig2012}{2012}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2015}{2015}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2017}{2017}), De Lisle,
Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).

Much of the theory and empirical data for character displacement between
the sexes is based on release from resource competition specifically
(e.g., Bolnick and Doebeli
(\protect\hyperlink{ref-BolnickandDoebeli2003}{2003}); Cooper, Gilman,
and Boughman (\protect\hyperlink{ref-Cooperetal2011}{2011}); Pfennig and
Pfennig (\protect\hyperlink{ref-PfennigandPfennig2012}{2012}); De Lisle
and Rowe (\protect\hyperlink{ref-DeLisleandRowe2015}{2015}); De Lisle,
Paiva, and Rowe (\protect\hyperlink{ref-DeLisleetal2018}{2018})).
However, release from predation might also drive the evolution of
increased sexual dimorphism because an absence of predators should
generate ecological opportunity (Reimchen and Nosil
(\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Parent and Crespi
(\protect\hyperlink{ref-ParentandCrespi2009}{2009}); Herrmann, Stroud,
and Losos (\protect\hyperlink{ref-Herrmannetal2021}{2021})).

Here, we tested the prediction that a release from predation results in
the evolution of increased sexual dimorphism using a well-preserved,
finely-resolved sequence of a fossil threespine stickleback fish
(Gasterosteus doryssus). The sequence in this depositional environment
is comprised of two lineages (Bell
(\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage I was a
low-armor form with zero to one dorsal spines and highly reduced
pelvises, on average. Lineage I lasted for at least 93,000 years before
it was replaced by a second lineage, suddenly, likely on the order of
years (Bell, Baumgartner, and Oslon
(\protect\hyperlink{ref-Belletal1985}{1985}); Stuart et al.~unpublished
data). Lineage II appeared in the depositional environment fully
armored, with complete pelvic girdles, two pelvic spines, and three
dorsal spines (Bell, Travis, and Blouw
(\protect\hyperlink{ref-Belletal2006}{2006}); Stuart, Travis, and Bell
(\protect\hyperlink{ref-Stuartetal2020}{2020})). It then immediately
began evolving adaptive reduction in its armor traits (Hunt, Bell, and
Travis (\protect\hyperlink{ref-Huntetal2008}{2008}); Stuart, Travis, and
Bell (\protect\hyperlink{ref-Stuartetal2020}{2020})) until it reached
the same low-armor state previously held by lineage I.

The observation of 93,000 years of low-armor stasis in lineage I (Bell,
Baumgartner, and Oslon (\protect\hyperlink{ref-Belletal1985}{1985})) and
the observation of rapid evolution of armor loss by lineage II, both
suggest that this depositional environment lacked piscivorous fish like
trout and other salmonids known to prey on modern threespine
stickleback. Armor presence in extant threespine stickleback
(Gasterosteus aculeatus) correlates strongly with the presence of
vertebrate piscivores; populations with less predation pressure
typically have less armor (Reimchen
(\protect\hyperlink{ref-Reimchen1994}{1994}), Reimchen and Nosil
(\protect\hyperlink{ref-ReimchenandNosil2004}{2004}); Bell et al.
(\protect\hyperlink{ref-Belletal1993}{1993}); Roesti et al.
(\protect\hyperlink{ref-Roestietal2023}{2023})). In our paleolake basin,
only three fossil trout have been found in the same section of rock that
has revealed \textgreater20,000 threespine stickleback fossils as well
as occasional killifish (Fundulus nevadensis) (Bell
(\protect\hyperlink{ref-Bell2009}{2009}); Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Thus, we reconstruct
an evolutionar history in which it is likely that lineage II migrated
from a nearby paleolake basin that had predators since it was armored
when it arrived (Bell (\protect\hyperlink{ref-Bell2009}{2009});
Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})). Lineage II then
experienced release from predators in the focal paleolake basin,
generating the initial conditions of evolutionary models predicting the
evolution of increased sexual dimorphism following ecological release.
We test this prediction here.

@@

For our study, we inferred sex in G. doryssus fossils by comparing
multivariate morphological trait data from lineage II samples to the
same multivariate trait set collected from multiple populations in the
closely-related, extant Threespine Stickleback species complex
(Gasterosteus aculeatus). Crucially, we determined G. aculeatus sex
directly via dissection and/or PCR genotyping. This enabled us to use
Multiple Imputation by Chained Equations (MICE; Buuren and
Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) to build a
predictive multiple imputation algorithm (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) based on the
multivariate morphology of G. aculeatus individuals of known sex. We
applied this algorithm to the fossil data to impute individual sex based
on morphology, treating fossil sex as a missing variable. With sex
assigned to fossil specimens, we fit a modified Ornstein-Uhlenbeck (OU)
(Uhlenbeck and Ornstein (\protect\hyperlink{ref-OUProcess}{1930})) model
using a Bayesian framework that accounts for uncertainty in sex
classification to test for evolution of sexual dimorphism in each trait
over \textasciitilde16,000 years of lineage II.

@@

We used Gasterosteus doryssus data that were previously reported by
Stuart, Travis, and Bell (\protect\hyperlink{ref-Stuartetal2020}{2020}),
Voje, Bell, and Stuart (\protect\hyperlink{ref-Vojeetal2022}{2022}), and
Siddiqui et al. (\protect\hyperlink{ref-Siddiquietal2024}{2024}).
Briefly, the data were collected from fossil Series K from Quarry D
(Cerasoni, Bell, and Stuart
(\protect\hyperlink{ref-Cerasonietal2024}{2024})), dug from an open pit
diatomite mine at 9.526° N, 119.094° W, near Hazen, Nevada, USA. Series
K consisted of 18 samples taken at \textasciitilde1000-year intervals,
and mean sample times span \textasciitilde16,363 years. Fish from series
K were measured for 16 ecomorphological traits related to armor,
swimming, and feeding (Table 1). Series K started at the previously
documented horizon when lineage I was replaced by lineage II. The tempo
and mode of lineage II armor reduction during this sequence suggests
adaptive evolution by natural selection (Hunt, Bell, and Travis
(\protect\hyperlink{ref-Huntetal2008}{2008})), and we focus on the
multivariate evolution of sexual dimorphism by this second lineage.

@@

To span the gamut of stickleback diversity for our predictive imputation
model, we sampled modern stickleback from lakes containing generalist
stickleback populations (Hendry et al.
(\protect\hyperlink{ref-Hendryetal2009}{2009}); Bolnick
(\protect\hyperlink{ref-Bolnick2011}{2011})) and from lakes containing
benthic-limnetic species pairs (Baumgartner, Bell, and Weinberg
(\protect\hyperlink{ref-Baumgartneretal1988}{1988}); Schluter and
McPhail (\protect\hyperlink{ref-SchluterandMcPhail1992}{1992})). The
generalist specimens used here were collected by YES in 2013 and were
previously described in Stuart et al.
(\protect\hyperlink{ref-Stuartetal2017}{2017}) (Table S1). These samples
were fixed in formalin, then stained for bone with Alizarin Red in 2013.
Benthic and limnetic specimens were kindly loaned by D. Schluter and S.
Blain at the University of British Columbia. The Schluter lab collected
benthic and limnetic individuals from Enos Lake in 1988 and from Emily
Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table
S1). The Enos specimens had been fixed whole in formalin and stored in
40\% isopropanol. The specimens from the other lakes were initially
preserved whole in 95\% ethanol in the field before being gradually
transferred to water then formalin in the lab and ultimately stored in
40\% isopropanol. In 2019, we stained these specimens for bone using
Alizarin Red. We next replicated fossil data collection (Table 1) on
these extant specimens. Standard length as well as pelvic-spine length
on each side were measured with calipers. We used a dissection
microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and
anal-fin rays. Right and left-side pelvic girdle lengths and
ectocoracoid lengths were measured from ventral photographs taken using
a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a
leveled Kaiser RS1 copy stand. Specimens were held in place for ventral
photographs using a small tabletop vise with an attached scale bar.
Lateral X-rays were used to measure dorsal spine length, number of
pterygiophores anterior to the pterygiophore holding the third spine,
length of the pterygiophore just anterior to the third spine, cleithrum
length, and pre-maxilla ascending branch length. We also counted
vertebrae from the X-rays: abdominal vertebrae were counted anterior to
the first vertebra with a haemal spine contacting an anal fin
pterygiophore. Caudal vertebrae were posterior, including the first
vertebra with the haemal spine contacting the anal fin pterygiophore
(following Aguirre, Walker, and Gideon
(\protect\hyperlink{ref-Aguirreetal2014}{2014})). X-rays were taken with
an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the
Field Museum of Natural History. Specimens were exposed at 35kV and 4mA.
Small fish were exposed for 7s, medium fish for 8s, and large fish for
10s. We developed the film and scanned individual images of each fish
using the B\&W Negatives setting on an Epson Perfection 4990 Photo
flatbed at 2400 dpi. Measurements from photographs and X-rays were taken
with FIJI (Schindelin et al.
(\protect\hyperlink{ref-Schindelinetal2012}{2012})) and its plugin
ObjectJ (\url{https://sils.fnwi.uva.nl/bcb/objectj/}).

@@

To check for outliers, we calculated within-group means and standard
deviations for each trait separately for K series fossil specimens
(pooled across samples) and for extant specimens (within generalist,
benthic, or limnetic categories). We noted trait values greater than 3.0
standard deviations from the mean as potential outliers. We deemed 3.0
s.d. to be a reasonable threshold for detecting errors without excluding
biologically relevant values. We checked whether these potential
outliers were a result of data entry and collection error and corrected
them if they were. We turned the remaining outlier trait values to NAs.
We size-corrected continuous traits only, as they varied with size,
unlike count traits that are fixed during early development. We
regressed each trait on standard length using a mixed-model regression,
pooling all specimens, following Stuart et al.
(\protect\hyperlink{ref-Stuartetal2017}{2017}). We appended the size
corrected data to the uncorrected trait data frame and used all of our
data (raw and size-corrected) to build the imputation model described
next.

@@

Briefly, We used multiple imputation (Little and Rubin
(\protect\hyperlink{ref-little2002statistical}{2002})) to impute the sex
of the fossils. Imputations were performed using the MICE (Buuren and
Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011})) algorithm and
implemented in R (Team (\protect\hyperlink{ref-R2022language}{2022}))
with M = 100 imputations. After imputing the sex, we fit a modified
Orenstein-Uhlenbeck (OU; Uhlenbeck and Ornstein
(\protect\hyperlink{ref-OUProcess}{1930})) model to each imputed data
set using a Bayesian framework to \{THIS NEEDS MORE\}. We then
quantified sexual dimorphism as the \{THIS NEEDS MORE; MAYBE WE CAN
WRITE TOGETHER\}. We describe the details of the imputation model and OU
modeling next.

@@

\textcolor{red}{AKHIL'S VALIDATION OF THE IMPUTATION MODEL GOES HERE}

We impute missing sex for the fossil data by sampling from the posterior
predictive distribution
\(P(\boldsymbol{W}_{fossil}|\boldsymbol{W}_{extant},\boldsymbol{Y})\)
using the multiple imputation by chained equations (MICE) algorithm
(Buuren and Groothuis-Oudshoorn (\protect\hyperlink{ref-MICE}{2011}))
with predictive mean matching. Traditionally, the choice for the number
of completed dats sets is a relatively small number such as \(M = 5\) or
\(M = 10\). However, Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}) recommend a larger number
of imputed data sets if users intend on performing Bayesian analysis
after imputation. Therefore, we imputed \(M = 100\) complet datasets.

\textcolor{red}{presumably this also imputes missing data in the other traits? We should probably say that explicitly. -YS}

We then pooled all of the draws from the posterior distribution across
all the imputed data sets to estimate the posterior distributions of our
parameters of interest (i.e., the trait values), rather than
usingRubin's combining rules, following Zhou and Reiter
(\protect\hyperlink{ref-ZhouReiter2010}{2010}). We used these posterior
distributions in our Bayesian analysis.

\hypertarget{bayesian-ou-modeling}{%
\subsection{Bayesian OU modeling}\label{bayesian-ou-modeling}}

For Bayesian modeling, we dropped the extant data and focused on
modeling the evolution of sexual dimorphism through time in the fossils.

\textcolor{red}{DOES THE NEXT SENTENCE CONTRADICT THE PREVIOUS ONE? BELOW, WE SAY THAT WE USED EACH IMPUTED DATA SET. ABOVE, WE SAY THAT WE POOLED ACROSS IMPUTED DATA SETS.-YS}
For a given imputed dataset, let \(W_{ij}\) be the imputed sex and
\(\boldsymbol{Y}_{ij}\) be the \(K \times 1\) vector of phenotypes for
stickleback fossil \(j\) at time \(t_i\) where \(i = 1, \ldots, T\) and
\(j = 1,\ldots,n_{t}\). Note, we are inputting only the fossil data into
our model; connecting to the previous section, \(W_{ij}\) and
\(\boldsymbol{Y}_{ij}\) can be interpreted as \(W_{fossil,ij}\) and
\(\boldsymbol{Y}_{fossil,ij}\). In addition, we denote \(Y_{K,ij}\), the
last variable in \(\boldsymbol{Y}_{ij}\), to be the standard length of
the fish, \begin{align}
{Y}_{K,ij} & \overset{iid}{\sim}\left\{\begin{array}{lll} \mathcal{N}(\mu_{K,ft_i},&\sigma_{K}^2), & W_{ij} = \text{Female} \\ \mathcal{N}(\mu_{K,mt_i},&\sigma_{K}^2), & W_{ij} = \text{Male} \end{array}\right..
\label{eq:stl}
\end{align}

@@

Because we are fitting a dataset with a stochastic structure on
phenotype means, and we are interested in the distribution of the
overall structure of these means, we analyze the data via a Bayesian
analysis \textless THIS LAST CLAUSE DOESN'T FOLLOW FROM THE BECAUSE
CLAUSE. BETTER EXPLAIN WHY BAYES IS GOOD HERE\textgreater. Bayesian data
analysis is also more naturally used when we have to impute data
(\textbf{CITATION}). To aid in the sampling procedure for the discrete
phenotypes, we declare
